$R^*$: A robust MCMC convergence diagnostic with uncertainty using gradient-boosted machines

### 2020-03-17 src/s_examples.Rmd
first draft of paper

```{r}
rm(list=ls())
library(tidyverse)
library(reshape2)
library(rstan)
library(latex2exp)
library(caret)
library(gbm)
library(mvtnorm)
options(mc.cores=4)
rstan_options(auto_write = TRUE)
```

@@

# Bivariate normal
```{r}
rmvrnorm2D <- function(n, mux, muy, sigmax, sigmay, rho){
  return(rmvnorm(n, c(mux, muy),
                 matrix(c(sigmax^2, sigmax * sigmay * rho,
                          sigmax * sigmay * rho, sigmay^2),
                        ncol = 2)))
}
# calling function
n <- 1000
chain1 <- rmvrnorm2D(n, 0, 0, 1, 1, 0.0) %>% 
  as.data.frame() %>% 
  mutate(chain=1)
chain2 <- rmvrnorm2D(n, 0, 0, 1, 1, 0.0) %>% 
  as.data.frame() %>% 
  mutate(chain=2)
chain3 <- rmvrnorm2D(n, 0, 0, 1, 1, 0.0) %>% 
  as.data.frame() %>% 
  mutate(chain=3)
chain4 <- rmvrnorm2D(n, 0, 0, 1, 1, 0.9) %>% 
  as.data.frame() %>% 
  mutate(chain=4)

chains_stacked <- rbind.data.frame(chain1, chain2, chain3, chain4) %>% 
  mutate(chain=as.factor(chain))

full_data <- chains_stacked
rand_samples <- sample(1:nrow(full_data), 0.7 * nrow(full_data))
training_data <- full_data[rand_samples, ]
testing_data <- full_data[-rand_samples, ]
gbmFit1 <- train(chain ~ ., data = training_data, 
                 method = "gbm",
                 trControl = trainControl(method = 'none'), 
                    tuneGrid = caretGrid, verbose=FALSE)

plda <- predict(object=gbmFit1, newdata=testing_data)
tibble(predicted=plda, actual=testing_data$chain) %>% 
  mutate(correct=if_else(predicted==actual, 1, 0)) %>% 
  summarise(mean(correct))
```

>>>>

```{r}
rm(list=ls())
library(tidyverse)
library(reshape2)
library(rstan)
library(latex2exp)
library(caret)
library(gbm)
library(mvtnorm)
options(mc.cores=4)
rstan_options(auto_write = TRUE)
source("monitornew.R")
```

@@

# Bivariate normal
```{r}
rmvrnorm2D <- function(n, mux, muy, sigmax, sigmay, rho){
  return(rmvnorm(n, c(mux, muy),
                 matrix(c(sigmax^2, sigmax * sigmay * rho,
                          sigmax * sigmay * rho, sigmay^2),
                        ncol = 2)))
}
# calling function
n <- 1000
chain1 <- rmvrnorm2D(n, 0, 0, 1, 1, 0.0) %>% 
  as.data.frame() %>% 
  mutate(chain=1,
         iter=seq_along(chain))
chain2 <- rmvrnorm2D(n, 0, 0, 1, 1, 0.0) %>% 
  as.data.frame() %>% 
  mutate(chain=2,
         iter=seq_along(chain))
chain3 <- rmvrnorm2D(n, 0, 0, 1, 1, 0.0) %>% 
  as.data.frame() %>% 
  mutate(chain=3,
         iter=seq_along(chain))
chain4 <- rmvrnorm2D(n, 0, 0, 1, 1, 0.9) %>% 
  as.data.frame() %>% 
  mutate(chain=4,
         iter=seq_along(chain))

chains_stacked <- rbind.data.frame(chain1, chain2, chain3, chain4) %>% 
  mutate(chain=as.factor(chain))
chains_stacked1 <- chains_stacked %>% 
  select(-iter)

full_data <- chains_stacked1
rand_samples <- sample(1:nrow(full_data), 0.7 * nrow(full_data))
training_data <- full_data[rand_samples, ]
testing_data <- full_data[-rand_samples, ]
gbmFit1 <- train(chain ~ ., data = training_data, 
                 method = "gbm",
                 trControl = trainControl(method = 'none'), 
                    tuneGrid = caretGrid, verbose=FALSE)

plda <- predict(object=gbmFit1, newdata=testing_data, type = "prob")
nsim <- 1000
mAccuracy <- matrix(nrow = nrow(plda),
                    ncol = nsim)
for(i in 1:nsim){
  for(j in 1:nrow(plda)){
    mAccuracy[j, i] <- if_else(which(rmultinom(1, 1, prob = plda[j, ])==1)==testing_data$chain[j], 1, 0)
  }
}

# 
# # use Vehtari et al. script to calculate Rhat
# chains_wide1 <- chains_stacked %>% 
#   pivot_wider(values_from = c("V1"), names_from = c("chain"), id_cols = "iter")
# rhat_rfun(as.matrix(chains_wide1[, 2:5]))
# chains_wide2 <- chains_stacked %>% 
#   pivot_wider(values_from = c("V2"), names_from = c("chain"), id_cols = "iter")
# rhat_rfun(as.matrix(chains_wide2[, 2:5]))
# 
# # Use Vehtari et al. to calculate ess
# ess_rfun(as.matrix(chains_wide1[, 2:5]))
# ess_rfun(as.matrix(chains_wide2[, 2:5]))
# ess_bulk(as.matrix(chains_wide1[, 2:5]))
# ess_bulk(as.matrix(chains_wide2[, 2:5]))
# ess_tail(as.matrix(chains_wide1[, 2:5]))
# ess_tail(as.matrix(chains_wide2[, 2:5]))
# ess_quantile(as.matrix(chains_wide1[, 2:5]), 0.95)
# ess_tail(as.matrix(chains_wide2[, 2:5]))

# Try monitor functionality -- split chains first
half_iter <- max(chains_stacked$iter) / 2
m_samples <- array(dim = c(half_iter,
                           2*n_distinct(chains_stacked$chain),
                           2))
k <- 1
chains_stacked_first <- chains_stacked %>% 
  filter(iter<=half_iter)
chains_stacked_second <- chains_stacked %>% 
  filter(iter>half_iter)

for(i in 1:half_iter){
  temp <- chains_stacked_first %>% 
    filter(iter==i)
  for(j in 1:4){
    temp1 <- temp %>%
      filter(chain==j)
    m_samples[i, j, ] <- as.numeric(temp1[1, 1:2])
    k <- k + 1
  }
}

for(i in 1:half_iter){
  temp <- chains_stacked_second %>% 
    filter(iter==(i + 500))
  for(j in 5:8){
    temp1 <- temp %>%
      filter(chain==(j - 4))
    m_samples[i, j, ] <- as.numeric(temp1[1, 1:2])
    k <- k + 1
  }
}

g <- 
  qplot(4*colMeans(mAccuracy)) +
  xlab(TeX("$R*$")) +
  xlim(0.9, NA) +
  geom_vline(xintercept = 1, linetype=2) +
  ylab("Count") +
  theme(text=element_text(size=14, colour="black"))
ggsave("../output/bivariate.pdf", g, width = 8, height = 6)
mon <- monitor(m_samples)

```

### 2020-03-17 write_up/ms.tex
first draft of paper

\begin{algorithm}[tb]
	\caption{$R^*$ calculation}
	\label{alg:R_star}
	\begin{algorithmic}
		\STATE Given chain-wise samples from the target, $\{X^{\{1\}},X^{\{2\}},...,X^{\{N\}}\}$ and a test set length, $S_\text{test}$:
		\FOR{$m=1$ to $N$}
		\STATE Create train and test sets by random-sampling (w/o replacement), $X^{\{m\}}\rightarrow\{X^{\{m\}}_\text{train},X^{\{m\}}_\text{test}\}$
		\ENDFOR
		\STATE Stack $X_\text{train} = (X^{\{1\}}_\text{train},X^{\{2\}}_\text{train},...,X^{\{N\}}_\text{train})^T$
		\STATE Stack $X_\text{test} = (X^{\{1\}}_\text{test},X^{\{2\}}_\text{test},...,X^{\{N\}}_\text{test})^T$
		\STATE Train ML model to classify chain id from any sample, $x$: $\text{ML}(x|X_\text{train}) \rightarrow c$
		\FOR{$s=1$ to $S_\text{test}$}
		\STATE Obtain test sample, $x^{\{s\}}=X_\text{test}(s)\in \mathbb{R}^K$
		\STATE Predict chain id, $c^{\{s\}} = \text{ML}(x^{\{s\}}|X_\text{train})$
		\STATE Compare with actual id, $c^s$: $a^{\{s\}}=\mathbbm{1}(c^{\{s\}}=c^s)$
		\ENDFOR
		\STATE Calculate predictive accuracy, $\bar{a} = \frac{1}{S_\text{test}} \sum_{s=1}^{S_\text{test}} a^{\{s\}}$
		\STATE Calculate ratio to null model accuracy, $R^* = \bar{a} / (1 / N) = N/\bar{a}$
		\RETURN $R^*$
	\end{algorithmic}
\end{algorithm}

@@

\begin{algorithm}[tb]
	\caption{Procedure to generate $I$ samples of $R^*$}
	\label{alg:R_star_uncertainty}
	\begin{algorithmic}
		\STATE Given test data $X_\text{test}$, number of chains $N$, number of iterations $I$, and fitted
		\STATE model, $\text{ML}(x|X_\text{train})\rightarrow(p_1,p_2,...,p_N)$:
		\FOR{$i=1$ to $I$}
		\FOR{$s=1$ to $S_\text{test}$}
		\STATE Obtain test sample, $x^{\{s\}}=X_\text{test}(s)\in \mathbb{R}^K$
		\STATE Predict chain id probabilities, $(p_1^{\{s\}},p_2^{\{s\}},...,p_N^{\{s\}})= \text{ML}(x^{\{s\}}|X_\text{train})$
		\STATE Draw a chain id, $c^{\{s\}} \sim \text{categorical}(p_1^{\{s\}},p_2^{\{s\}},...,p_N^{\{s\}})$
		\STATE Compare with actual id, $c^s$: $a^{\{s\}}=\mathbbm{1}(c^{\{s\}}=c^s)$
		\ENDFOR
		\STATE Calculate predictive accuracy, $\bar{a} = \frac{1}{S_\text{test}} \sum_{s=1}^{S_\text{test}} a^{\{s\}}$
		\STATE Calculate ratio to null model accuracy, $R^{*i} = \bar{a} / (1 / N) = N/\bar{a}$
		\ENDFOR
		\RETURN $(R^{*1},R^{*2},...,R^{*I})$
	\end{algorithmic}
\end{algorithm}

\section{Results}\label{sec:results}
To illustrate the versatility of $R^*$, we use examples from \cite{vehtari2019rank} with one additional example: the multivariate normal model described in \cite{hoffman2014no}.

@@

\subsection{Diagnosing unconvergence in joint distribution: bivariate normal model}\label{sec:bivariate_normal}

>>>>

\begin{algorithm}[tb]
	\caption{$R^*$ calculation}
	\label{alg:R_star}
	\begin{algorithmic}
		\STATE Given chain-wise samples from the target, $\{X^{\{1\}},X^{\{2\}},...,X^{\{N\}}\}$ and a test set length, $S_\text{test}$:
		\FOR{$m=1$ to $N$}
		\STATE Create train and test sets by random-sampling (w/o replacement), $X^{\{m\}}\rightarrow\{X^{\{m\}}_\text{train},X^{\{m\}}_\text{test}\}$
		\ENDFOR
		\STATE Stack $X_\text{train} = (X^{\{1\}}_\text{train},X^{\{2\}}_\text{train},...,X^{\{N\}}_\text{train})^T$
		\STATE Stack $X_\text{test} = (X^{\{1\}}_\text{test},X^{\{2\}}_\text{test},...,X^{\{N\}}_\text{test})^T$
		\STATE Train ML model to classify chain id from any sample, $x$: $\text{ML}(x|X_\text{train}) \rightarrow c$
		\FOR{$s=1$ to $S_\text{test}$}
		\STATE Obtain test sample, $x^{\{s\}}=X_\text{test}(s)\in \mathbb{R}^K$
		\STATE Predict chain id, $c^{\{s\}} = \text{ML}(x^{\{s\}}|X_\text{train})$
		\STATE Compare with actual id, $c^s$: $a^{\{s\}}=\mathbbm{1}(c^{\{s\}}=c^s)$
		\ENDFOR
		\STATE Calculate predictive accuracy, $\bar{a} = \frac{1}{S_\text{test}} \sum_{s=1}^{S_\text{test}} a^{\{s\}}$
		\STATE Calculate ratio to null model accuracy, $R^* = \bar{a} / (1 / N) = N \bar{a}$
		\RETURN $R^*$
	\end{algorithmic}
\end{algorithm}

@@

\begin{algorithm}[tb]
	\caption{Procedure to generate $I$ samples of $R^*$}
	\label{alg:R_star_uncertainty}
	\begin{algorithmic}
		\STATE Given test data $X_\text{test}$, number of chains $N$, number of iterations $I$, and fitted
		\STATE model, $\text{ML}(x|X_\text{train})\rightarrow(p_1,p_2,...,p_N)$:
		\FOR{$i=1$ to $I$}
		\FOR{$s=1$ to $S_\text{test}$}
		\STATE Obtain test sample, $x^{\{s\}}=X_\text{test}(s)\in \mathbb{R}^K$
		\STATE Predict chain id probabilities, $(p_1^{\{s\}},p_2^{\{s\}},...,p_N^{\{s\}})= \text{ML}(x^{\{s\}}|X_\text{train})$
		\STATE Draw a chain id, $c^{\{s\}} \sim \text{categorical}(p_1^{\{s\}},p_2^{\{s\}},...,p_N^{\{s\}})$
		\STATE Compare with actual id, $c^s$: $a^{\{s\}}=\mathbbm{1}(c^{\{s\}}=c^s)$
		\ENDFOR
		\STATE Calculate predictive accuracy, $\bar{a} = \frac{1}{S_\text{test}} \sum_{s=1}^{S_\text{test}} a^{\{s\}}$
		\STATE Calculate ratio to null model accuracy, $R^{*i} = \bar{a} = N\bar{a}$
		\ENDFOR
		\RETURN $(R^{*1},R^{*2},...,R^{*I})$
	\end{algorithmic}
\end{algorithm}

\section{Results}\label{sec:results}
To illustrate the versatility of $R^*$, we use examples from \cite{vehtari2019rank} two additional examples based on a multivariate normal model.

@@

\subsection{Diagnosing convergence in joint distributions: bivariate normal model}\label{sec:bivariate_normal}
To illustrate how $R^*$ can diagnose convergence issues in the joint target distribution, we consider a bivariate normal density. In all four chains, we use independent sampling to generate 1000 draws from bivariate normal densities; in three of these chains, the covariance matrix is an identity matrix; in one chain, the covariance matrix also has unit diagonal terms but has off-diagonal terms of 0.9, indicating strong covariance between the two dimensions. By construction, all chains target the same marginal distribution in each dimension but the fourth chain has a different joint distribution.

First, we use the code provided in \cite{vehtari2019rank} to calculate rank-normalised $\hat{R}$ and two different ESS measures, aiming to capture how well certain regions of the posterior have been explored: these are known as bulk ESS and tail ESS. In all cases, the various quantities were calculated based on chains split into halves. For both dimensions, the two ESS measures were above 3700 and $\hat{R}<1.001$, indicating no issues with convergence.

Next, we estimate the $R^*$ distribution which is shown in Fig. \ref{fig:bivariate}. The mean of this distribution is 1.18 and all the samples are above 1 -- indicating that the sampling distribution has not converged. By taking account of all the information in the chains,

\begin{figure}[h]
	\centerline{\includegraphics[width=0.6\textwidth]{../output/bivariate.pdf}}
	\caption{\textbf{Bivariate normal example.} The distribution for $R^*$ across 1000 samples. See the R markdown file on the Github repository for code to reproduce this example.}
	\label{fig:bivariate}
\end{figure}
